Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-10-31 rmd/manuscript.Rmd
edit

---
title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  #pdf_document:
  #  latex_engine: xelatex
  word_document:
    reference_docx: word_format.docx
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: ecosystems.csl
header-includes:
  \usepackage{textgreek}
---

@@

We employed a discrete-time model to describe tri-trophic dynamics in each habitat patch. A discrete model is suitable for organisms with predictable reproduction cycles, a common life history for many lotic species, including macroinvertebrates and fish. Our simulation model consists of three species communities; a basal species $B$ with positive logistic growth up to a carrying capacity, $K$; a primary consumer species or intraguild prey, $C$, which grows based on the consumption of $B$; and an intraguild predator $P$, which can consume both $B$ and $C$ (**Figure 1C**). In our model, the realized species abundance $A_{i,x}(t+1)$ for species i (either $B,C,~or~P$) at patch $x$ and time $t+1$ is expressed as a composite of within-patch trophic interactions, immigration, and emigration:

@@

where $\bar{A}_{i,x}(t+1)$ is the expected species abundance after within-patch trophic interactions, $I_{i,x}(t)$ the expected number of immigrants into patch $x$, and $E_{i,x}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.

***Internal dynamics.*** Each time step starts with internal dynamics. The basal species is exposed to predation as well as other sources of mortality:

$$
A_{B,x}'(t) = s_{0,B}[1-z_{B,x}(t)]A_{B,x}(t)
$$

where $s_{0,i}$ is the background survival probability (density independent) and $z_{B,x}(t)$ the proportional mortality caused by predation. Let $W_{B,x}(t)$ be the total number of basal species consumed by $C$ and $P$. The predator-induced mortality $z_{B,x}(t)$, the sum of proportional predation by $C$ ($z_{BC,x}(t)$) and $P$ ($z_{BP,x}(t)$), is then:

$$
z_{B,x}(t) = z_{BC,x}(t) + z_{BP,x}(t) =\left\{\begin{aligned}&\frac{W_{B,x}(t)}{A_{B,x}(t)} &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&1 &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

and

$$
W_{B,x}(t) = W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t)
$$

@@

$$
\hat{A}_{B,x}(t) = \frac{r_{b} A_{B,x}'(t)}{1 + \frac{r_b - 1}{K_{x}}A_{B,x}'(t)}
$$

@@

The population growth of the consumer is controlled by conversion efficiency, number of prey captured, background mortality, and predation:

$$
\hat{A}_{C,x}(t) =\left\{\begin{aligned}&s_{0,C} (1 - z_{CP,x}) e_{BC}W_{BC,x}(t) &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&s_{0,C} (1 - z_{CP,x}) e_{BC} \left( \frac{W_{BC,x}(t)}{W_{B,x}(t)} A_{B,x} \right) &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

@@

The population of the predator is fueled by two prey species:

$$
\hat{A}_{P,x}(t) =\left\{\begin{aligned}&s_{0,P} \left[e_{BP} \delta_{x}(t) W_{BP,x}(t) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\&s_{0,P} \left[e_{BP} \delta_{x}(t)\left(\frac{W_{BP,x}(t)}{W_{B,x}(t)}A_{B,x}\right) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) \right] &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

@@

$$
\bar{A}_{i,x}(t) = [1 - \psi(t) m_{x}] \hat{A}_{i,x}(t)
$$

@@

where $d_{xy}$ is the network distance between patch $x$ and $y$ (shortest path), and $\theta$ indicates the inverse of mean dispersal distance (i.e., larger numbers mean shorter dispersal distance). The immigration potential at patch $x$, $\sum_{y, y \ne x}E_{i,y}(t)exp(-\theta d_{xy})$, is scaled by the sum of immigration potential across patches so that $\sum_{x} \xi_{i,x}=1$. The expected number of immigrants into patch $x$ is calculated as $I_{i,x}(t)=\xi_{i,x}(t) \sum_{x}E_{i,x}(t)$.

>>>>

---
title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
  #word_document:
  #  reference_docx: word_format.docx
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: ecosystems.csl
header-includes:
  \usepackage{textgreek}
---

@@

We employed a discrete-time model to describe tri-trophic dynamics in each habitat patch. A discrete model is suitable for organisms with predictable reproduction cycles, a common life history for many lotic species, including macroinvertebrates and fish. Our simulation model consists of three species; a basal species $B$ with positive logistic growth up to a carrying capacity, $K$; a primary consumer species or intraguild prey, $C$, which grows based on the consumption of $B$; and an intraguild predator $P$, which can consume both $B$ and $C$ (**Figure 1C**). In our model, the realized species abundance $A_{i,x}(t+1)$ for species i (either $B,C,~or~P$) at patch $x$ and time $t+1$ is expressed as a composite of within-patch trophic interactions, immigration, and emigration:

@@

where $\bar{A}_{i,x}(t)$ is the expected species abundance after within-patch trophic interactions, $I_{i,x}(t)$ the expected number of immigrants into patch $x$, and $E_{i,x}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.

***Internal dynamics.*** Each time step starts with internal dynamics. We accounted for background mortality, predator-induced mortality, and disturbance-induced mortality before the density-dependent growth occurs:

$$
\begin{aligned}
A'_{i,x} &= s_{0,i}~\beta_{i,x}(t)~A_{i,x}(t)\\
\bar{A}_{i,x}(t) &= g_i(A'_{i,x})~A'_{i,x}~\gamma_{x}(t)
\end{aligned}
$$

where $s_{0,i}$ is the background survival probability (density independent), $\beta_{i,x}(t)$ the survival probability after predation, and $\gamma_{x}(t)$ the survival probability after stochastic disturbance. The function $g_i(\centerdot)$ defines the form of the density-dependent recruitment for species $i$.

For the basal species, we considered the predation impacts by consumer and predator species:

$$
\beta_{B,x}(t) = 1 - [z_{BC,x}(t) + z_{BP,x}(t)]
$$
where $z_{BC,x}(t)$ and $z_{BP,x}(t)$ are the proportional mortality of basal species caused by consumer and predator species, respectively. The total mortality is calculated as the proportion of total individuals consumed ($W_{B,x}(t)=W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t)$) relative to the current abundance of the basal species ($A_{B,x}(t)$):

$$
z_{BC,x}(t) + z_{BP,x}(t) =\left\{\begin{aligned}&\frac{W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t)}{A_{B,x}(t)} &&\text{if $W_{B,x}(t) < A_{B,x}(t)$}\\&1 &&\text{if $W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

@@

$$
g_B(A'_{B,x}) = \frac{r_{b}} {1 + \frac{r_b - 1}{K_{x}}A_{B,x}'}
$$

@@

For the consumer species, the predation-induced mortality is the proportion of individuals consumed by the predator species:

$$
\beta_{CP,x}(t) = 1 - z_{CP,x} = 1 - \frac{W_{CP,x}}{A_{C,x}(t)}
$$

The population growth of the consumer is controlled by conversion efficiency and the number of prey captured:

$$
g_C(A'_{C,x}) =
\left\{
\begin{aligned}
&e_{BC}W_{BC,x}(t) &&\text{if $W_{B,x}(t) < A_{B,x}(t)$}\\
&e_{BC} \left( \frac{W_{BC,x}(t)}{W_{B,x}(t)} A_{B,x} \right) &&\text{if $W_{B,x}(t) > A_{B,x}(t)$}\\\end{aligned}\right.
$$

@@

The predator species is not exposed to predation, thus $\beta_{P,x}(t) = 1$. The population growth of the predator considers two prey species:

$$
g_P(A'_{P,x}) =
\left\{
\begin{aligned}
&e_{BP} \delta_{x}(t) W_{BP,x}(t) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) &&\text{if $W_{B,x}(t) < A_{B,x}(t)$}\\
&e_{BP} \delta_{x}(t)\left(\frac{W_{BP,x}(t)}{W_{B,x}(t)}A_{B,x}\right) + e_{CP} (1 - \delta_{x}(t)) W_{CP,x}(t) &&\text{if $W_{B,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
$$

@@

$$
\gamma_{i,x}(t) = [1 - \psi(t) m_{x}]
$$

@@

where $d_{xy}$ is the network distance between patch $x$ and $y$ (shortest path), and $\theta$ indicates the inverse of mean dispersal distance (i.e., larger numbers mean shorter dispersal distance). The immigration potential at patch $x$, $\sum_{y, y \ne x}E_{i,y}(t)\exp(-\theta d_{xy})$, is scaled by the sum of immigration potential across patches so that $\sum_{x} \xi_{i,x}=1$. The expected number of immigrants into patch $x$ is calculated as $I_{i,x}(t)=\xi_{i,x}(t) \sum_{x}E_{i,x}(t)$.
